---
title: "Getting Started with watershedtools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with watershedtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(watershedtools)
```

## Overview

**watershedtools** provides a focused set of hydrologic analysis utilities
built on top of the NHDPlus data model and the `sf` spatial framework.
Core capabilities include:
- **HUC validation** — check and manipulate hydrologic unit codes
- **Watershed delineation** — retrieve drainage basin boundaries via NLDI
- **Stream tracing** — walk upstream or downstream through a flowline network
- **Pour point snapping** — move points onto the nearest stream reach
- **Spatial overlay** — join features (sites, permits, etc.) to watershed polygons

The package ships with sample data for **Elkhorn Creek, Kentucky** — a
real watershed in the Kentucky River basin (HUC8 05100205) — so every
example can run without an internet connection.

## Installation

```{r, eval = FALSE}
# From CRAN (when available)
install.packages("watershedtools")

# Development version
# install.packages("pak")
pak::pak("chrislyonsKY/watershedtools")
```

## Loading Sample Data

The package bundles three GeoJSON files covering Elkhorn Creek, KY:

```{r load-data}
# NHDPlus-style flowline network (7 segments)
flowlines <- sf::st_read(
  system.file("extdata", "elkhorn_flowlines.geojson", package = "watershedtools"),
  quiet = TRUE
)

# Watershed basin boundary (HUC8 05100205)
basin <- sf::st_read(
  system.file("extdata", "elkhorn_basin.geojson", package = "watershedtools"),
  quiet = TRUE
)

# USGS gages and KPDES monitoring sites
sites <- sf::st_read(
  system.file("extdata", "elkhorn_sites.geojson", package = "watershedtools"),
  quiet = TRUE
)
```

## 1 — Validate HUC Codes

HUC codes must be stored as character strings to preserve leading zeros.
`validate_huc()` checks length and digit content:

```{r huc-validate}
# Kentucky watershed HUCs
hucs <- c("05", "0510", "05100205", "051002050101")
validate_huc(hucs)

# Common mistakes
validate_huc(c("5", "abc", "0510020"))
```

Use `huc_level()` and `parent_huc()` for hierarchy navigation:

```{r huc-hierarchy}
huc_level(c("05", "0510", "05100205"))

# Get the HUC4 parent of a HUC8
parent_huc("05100205", level = 4)
```

## 2 — Trace Upstream and Downstream

The tracing functions walk through a local flowline network using the
NHDPlus `hydroseq` / `dnhydroseq` topology:

```{r trace}
# Everything upstream of Elkhorn Creek main stem (COMID 9734460)
upstream <- trace_upstream(flowlines, comid = 9734460)
nrow(upstream)

# Follow North Elkhorn Creek downstream to the mouth
downstream <- trace_downstream(flowlines, comid = 9734444)
downstream$gnis_name
```

```{r plot-trace, fig.width = 6, fig.height = 4}
plot(sf::st_geometry(flowlines), col = "grey70", lwd = 1,
     main = "Downstream from N. Elkhorn headwater")
plot(sf::st_geometry(downstream), col = "tomato", lwd = 2.5, add = TRUE)
```

## 3 — Snap a Pour Point

Place a point near a stream and snap it onto the nearest flowline:

```{r snap}
pt <- sf::st_sf(
  geometry = sf::st_sfc(sf::st_point(c(-84.68, 38.155)), crs = 4326)
)
snapped <- snap_pour_point(pt, flowlines)
snapped[, c("comid", "gnis_name", "snap_distance")]
```

```{r plot-snap, fig.width = 6, fig.height = 4}
plot(sf::st_geometry(flowlines), col = "steelblue", lwd = 2,
     main = "Pour point snapping")
plot(sf::st_geometry(pt), add = TRUE, col = "red", pch = 16, cex = 1.5)
plot(sf::st_geometry(snapped), add = TRUE, col = "green3", pch = 16, cex = 1.5)
legend("topright", legend = c("Original", "Snapped"),
       col = c("red", "green3"), pch = 16)
```

## 4 — Overlay Features with Watersheds

Determine which monitoring sites fall inside the Elkhorn Creek basin:

```{r overlay}
result <- overlay_features(sites, basin)
result[, c("site_id", "name", "huc8")]
```

## 5 — Delineate a Watershed (Network)

`get_watershed()` calls the NLDI web service to retrieve a basin boundary.
This requires the `nhdplusTools` package and an internet connection:

```{r watershed-delineate, eval = FALSE}
# Delineate the watershed for Elkhorn Creek at Frankfort
basin_live <- get_watershed(comid = 9734460)
plot(sf::st_geometry(basin_live), col = "lightblue", border = "steelblue")
```

## Real-World Data Sources

The `examples/` directory in this repository includes scripts demonstrating
integration with production data sources:

| Script | Data Source | Description |
|--------|-------------|-------------|
| `ky_watershed_analysis.R` | NHDPlus / NLDI | Full watershed analysis for Elkhorn Creek |
| `usgs_gage_delineation.R` | USGS NWIS | Delineate watersheds for Kentucky USGS gages |
| `permit_watershed_overlay.R` | KPDES / EPA ECHO | Overlay discharge permits with HUC boundaries |

See also the [USGS National Hydrography page](https://www.usgs.gov/national-hydrography)
and the [NLDI API](https://labs.waterdata.usgs.gov/api/nldi/) for the
upstream data services.
